Improving the portuguese hyphenation rules

### 2023-07-14 article/article.tex
fix #19
fix issue #24
fix #26

@@

Merely stating that an orthographic system is guided by phonological issues
does not necessarily mean that its hyphenation rules directly reflect the way a
word is pronounced -- this is because Portuguese does not have a direct
letter-sound correspondence, and vice versa.  The orthographic system operates
according to its own specific rules. For instance, although consonant clusters that
produce a single sound (i.e., digraphs) are pronounced within a single syllable
in words like \emph{chave}, \emph{ilha}, \emph{shampoo}, \emph{carro}, and
\emph{pássaro}, their representation in writing determines how they are
separated.  Specifically, different consonants within a digraph must remain
together, while different consonants are separated. As a result, we have
\emph{cha-ve}, \emph{i-lha}, and \emph{sham-poo}, but \emph{car-ro}.

@@

\begin{enumerate}
\item regra: \texttt{p2si}, \texttt{p2sí} -- psicologia, psíquico\\
exceção: \texttt{p3sia}, \texttt{p3sin} -- epilepsia, rodopsina

\item regra: \texttt{t2c} -- tchau%, tcheco

\item regra: \texttt{p2neu} -- pneumonia

\item regra: \texttt{.g2no}, \texttt{.g2nó} -- gnomo, gnóstico

\item regra: \texttt{.p2t} -- ptose%, pterossauro

\item regra: \texttt{su2b3r}, \texttt{su2b3l} -- sublunar, subrotina
exceção: \texttt{su3b4li1nh}, \texttt{su3b4li1ma}, \texttt{su3b4li1me}, \texttt{su3b4li1mid} -- sublinhar, sublimar, sublime, sublimidade

\item regra: \texttt{1sô} -- subsônico

\item regra: \texttt{1p2seu1d} -- pseudônimo

\item regra: \texttt{c2za} -- czar

\item regra: \texttt{.s2} -- stalinismo

\item regra: \texttt{.t2} -- tsunami

\item regra: \texttt{a1â}, \texttt{a1ã}, \texttt{a1é}, \texttt{a1í}, \texttt{a1ó}, \texttt{a1ô}, \texttt{a1ú}, \texttt{e1á}, \texttt{e1â}, \texttt{e1ã}, \texttt{e1ê}, \texttt{e1í}, \texttt{e1ó}, \texttt{e1ô}, \texttt{e1ú}, \texttt{i1á}, \texttt{i1ã}, \texttt{i1ã}, \texttt{i1é}, \texttt{i1ó}, \texttt{i1ú}, \texttt{o1á}, \texttt{o1ã}, \texttt{o1é}, \texttt{o1ê}, \texttt{o1í}, \texttt{o1ó}, \texttt{u1á}, \texttt{u1ã}, \texttt{u1é}, \texttt{u1ê}, \texttt{u1í} --  abraâmico, abraão, aéreo, país, caótico, faraônico, saúde, balneário, oceânico, campeã, veêm, veículo, teórico, napoleônico, conteúdo, diário, região, soviético, periódico, viúva, razoável, joão, poético, boêmia, heroísmo, alcoólico, usuário, itapuã, suécia, cauê, suíça \\
exceção: \texttt{1gu2é}, \texttt{1gu2ê}, \texttt{1gu2í}, \texttt{1qu2á}, \texttt{1qu2é}, \texttt{1qu2ê}, \texttt{1qu2í} -- alguém, português, linguística, aquático, inquérito, sequência, química

\item regra: \texttt{a1ir.}, \texttt{u1ir.} -- sair, diminuir

\item regra: \texttt{í1a} -- baía

\item regra: \texttt{1lô} -- camelô

\item regra: \texttt{1cô} -- recôncavo

\item regra: \texttt{e1in1c}, \texttt{e1in1f}, \texttt{e1in1g}, \texttt{e1in1s}, \texttt{e1in1t}, \texttt{e1in1v} -- reincidência, reinfecção, reingressa, reinserção, reintegração, reinventar
exceção: \texttt{e1in2st}, \texttt{e1in2sc} -- reinstalado, reinscrever

\item regra: \texttt{o1in} -- coincidência, agroindustrial, bioinformática, amendoins, gastrointestinal
exceção: \texttt{o1ina} -- boina

\item regra: \texttt{a1ind}, \texttt{a1i1nh} -- ainda, rainha

\item regra: \texttt{o1i1nh} -- moinho

\item regra: \texttt{1bô}, \texttt{1dô}, \texttt{1fô}, \texttt{1gô}, \texttt{1pô}, \texttt{1mô}, \texttt{1nô}, \texttt{1rô}, \texttt{1tô}, \texttt{1vô}, \texttt{1xô}, \texttt{1zô} -- robô, judô,  telefônica, xangô, capô, sumô, econômico, tarô, chatô, vovô, saxônia, amazônia

\item regra: \texttt{e1im} -- reimpresso, coimbra, coimperador
exceção: \texttt{que2i}, \texttt{te2im} -- queimar, queimadura, teimosia, tirateima

\item regra: \texttt{au1i1c}, \texttt{du1i1c}, \texttt{u1i1ç}, \texttt{u1i1d}, \texttt{cu2i}, \texttt{dru2i}, \texttt{flu2id}, \texttt{bu1i1n}, \texttt{cu1i1n}, \texttt{fu1i1n}, \texttt{nu1i1n}, \texttt{ru1i1na}, \texttt{ru1i1no}, \texttt{su1i1ti}, \texttt{tu1i1ti}, \texttt{u1iz} -- cacauicultor, sanduicheira, constituição, continuidade, cuidador, druida, fluido, tabuinha, picuinha, fuinha, genuinamente, arruinado, arruinou, jesuitismo, intuitivo, juizado 
exceção: \texttt{cu3i1da1de}, \texttt{bu1i2n1d}, \texttt{bu1i2n1t}  -- acuidade, contribuindo, contribuinte

\item regra: \texttt{.m2n} -- mnemônico

\item regra: \texttt{1gu4á}, \texttt{1gu4ã}, \texttt{1qu4ã} -- jaraguá, saguão, quão

\item regra: \texttt{1çô} -- maçônico

\item regra: \texttt{i1ur} -- diurno

\item regra: \texttt{í1o} -- íon

\item regra: \texttt{pro1i1b} -- proibição

\item regra: \texttt{i1un} -- triunfar

%\item regra: \texttt{u3ó} -- fluór

\item regra: \texttt{1qu4ó}, \texttt{1qu4â} -- quórum, equânime

\item regra: \texttt{é3o} -- alvéolo

\item regra: \texttt{i4a.}, \texttt{i4e.}, \texttt{i4o.}, \texttt{o4a.}, \texttt{u4a.} -- economia, espécie, vazio, destoa, institua

\item regra: \texttt{z1z}, \texttt{p1p}, \texttt{z2z.} -- pizza, shopping, jazz

% As sequências gráficas ia, ie, io, oa, ua quando em posição
% final átona, que são normalmente pronunciadas como ditongos
% crescentes, mas que podem corresponder foneticamente a duas
% vogais (hiato), sobretudo quando pronunciadas pausadamente:
% vitó-ria, espé-cie, exercí-cio, nó-doa, lé-gua, tê-nue, ambí-guo.
\end{enumerate}

>>>>

\newcounter{numberRulesGroups}

@@

Merely stating that an orthographic system is guided by phonological issues
does not necessarily mean that its hyphenation rules directly reflect the way a
word is pronounced -- this is because Portuguese does not have a direct
letter-sound correspondence, and vice versa.  The orthographic system operates
according to its own specific rules. For instance, although consonant clusters that
produce a single sound (i.e., digraphs) are pronounced within a single syllable
in words like \emph{chave}, \emph{ilha}, \emph{shampoo}, \emph{carro}, and
\emph{pássaro}, their representation in writing determines how they are
separated.  Specifically, different consonants within a digraph must remain
together, while identical consonants are separated. As a result, we have
\emph{cha-ve}, \emph{i-lha}, and \emph{sham-poo}, but \emph{car-ro}.

@@

\begin{enumerate}
\item\label{rulegrp_psi} regra: \texttt{p2si}, \texttt{p2sí} -- psicologia, psíquico\\
exceção: \texttt{p3sia}, \texttt{p3sin} -- epilepsia, rodopsina

\item\label{rulegrp_tc} regra: \texttt{t2c} -- tchau%, tcheco

\item\label{rulegrp_pneu} regra: \texttt{p2neu} -- pneumonia

\item\label{rulegrp_gno} regra: \texttt{.g2no}, \texttt{.g2nó} -- gnomo, gnóstico

\item\label{rulegrp_pt} regra: \texttt{.p2t} -- ptose%, pterossauro

\item\label{rulegrp_cza} regra: \texttt{c2za} -- czar

\item\label{rulegrp_s2} regra: \texttt{.s2} -- stalinismo

\item\label{rulegrp_t2} regra: \texttt{.t2} -- tsunami

\item\label{rulegrp_mn} regra: \texttt{.m2n} -- mnemônico

\item\label{rulegrp_zz} regra: \texttt{z1z}, \texttt{p1p}, \texttt{z2z.} -- pizza, shopping, jazz

\item\label{rulegrp_sub} regra: \texttt{su2b3r}, \texttt{su2b3l} -- sublunar, subrotina
exceção: \texttt{su3b4li1nh}, \texttt{su3b4li1ma}, \texttt{su3b4li1me}, \texttt{su3b4li1mid} -- sublinhar, sublimar, sublime, sublimidade

\item\label{rulegrp_so} regra: \texttt{1sô} -- subsônico

\item\label{rulegrp_pseu} regra: \texttt{1p2seu1d} -- pseudônimo

\item\label{rulegrp_ein} regra: \texttt{e1in1c}, \texttt{e1in1f}, \texttt{e1in1g}, \texttt{e1in1s}, \texttt{e1in1t}, \texttt{e1in1v} -- reincidência, reinfecção, reingressa, reinserção, reintegração, reinventar
exceção: \texttt{e1in2st}, \texttt{e1in2sc} -- reinstalado, reinscrever

\item\label{rulegrp_oin} regra: \texttt{o1in} -- coincidência, agroindustrial, bioinformática, amendoins, gastrointestinal
exceção: \texttt{o1ina} -- boina

\item\label{rulegrp_eim} regra: \texttt{e1im} -- reimpresso, coimbra, coimperador
exceção: \texttt{que2i}, \texttt{te2im} -- queimar, queimadura, teimosia, tirateima

\item\label{rulegrp_aa} regra: \texttt{a1â}, \texttt{a1ã}, \texttt{a1é}, \texttt{a1í}, \texttt{a1ó}, \texttt{a1ô}, \texttt{a1ú}, \texttt{e1á}, \texttt{e1â}, \texttt{e1ã}, \texttt{e1ê}, \texttt{e1í}, \texttt{e1ó}, \texttt{e1ô}, \texttt{e1ú}, \texttt{i1á}, \texttt{i1ã}, \texttt{i1ã}, \texttt{i1é}, \texttt{i1ó}, \texttt{i1ú}, \texttt{o1á}, \texttt{o1ã}, \texttt{o1é}, \texttt{o1ê}, \texttt{o1í}, \texttt{o1ó}, \texttt{u1á}, \texttt{u1ã}, \texttt{u1é}, \texttt{u1ê}, \texttt{u1í} --  abraâmico, abraão, aéreo, país, caótico, faraônico, saúde, balneário, oceânico, campeã, veêm, veículo, teórico, napoleônico, conteúdo, diário, região, soviético, periódico, viúva, razoável, joão, poético, boêmia, heroísmo, alcoólico, usuário, itapuã, suécia, cauê, suíça \\
exceção: \texttt{1gu2é}, \texttt{1gu2ê}, \texttt{1gu2í}, \texttt{1qu2á}, \texttt{1qu2é}, \texttt{1qu2ê}, \texttt{1qu2í} -- alguém, português, linguística, aquático, inquérito, sequência, química

\item\label{rulegrp_air} regra: \texttt{a1ir.}, \texttt{u1ir.} -- sair, diminuir

\item\label{rulegrp_ia} regra: \texttt{í1a} -- baía

\item\label{rulegrp_ain} regra: \texttt{a1ind}, \texttt{a1i1nh} -- ainda, rainha

\item\label{rulegrp_oin} regra: \texttt{o1i1nh} -- moinho

\item\label{rulegrp_aui} regra: \texttt{au1i1c}, \texttt{du1i1c}, \texttt{u1i1ç}, \texttt{u1i1d}, \texttt{cu2i}, \texttt{dru2i}, \texttt{flu2id}, \texttt{bu1i1n}, \texttt{cu1i1n}, \texttt{fu1i1n}, \texttt{nu1i1n}, \texttt{ru1i1na}, \texttt{ru1i1no}, \texttt{su1i1ti}, \texttt{tu1i1ti}, \texttt{u1iz} -- cacauicultor, sanduicheira, constituição, continuidade, cuidador, druida, fluido, tabuinha, picuinha, fuinha, genuinamente, arruinado, arruinou, jesuitismo, intuitivo, juizado 
exceção: \texttt{cu3i1da1de}, \texttt{bu1i2n1d}, \texttt{bu1i2n1t}  -- acuidade, contribuindo, contribuinte

\item\label{rulegrp_gua} regra: \texttt{1gu4á}, \texttt{1gu4ã}, \texttt{1qu4ã} -- jaraguá, saguão, quão

\item\label{rulegrp_iur} regra: \texttt{i1ur} -- diurno

\item\label{rulegrp_io} regra: \texttt{í1o} -- íon

\item\label{rulegrp_proi} regra: \texttt{pro1i1b} -- proibição

\item\label{rulegrp_iun} regra: \texttt{i1un} -- triunfar

\item\label{rulegrp_quo} regra: \texttt{1qu4ó}, \texttt{1qu4â} -- quórum, equânime

%\item\label{rulegrp_uo} regra: \texttt{u3ó} -- fluór

\item\label{rulegrp_eo} regra: \texttt{é3o} -- alvéolo

\item\label{rulegrp_ia} regra: \texttt{i4a.}, \texttt{i4e.}, \texttt{i4o.}, \texttt{o4a.}, \texttt{u4a.} -- economia, espécie, vazio, destoa, institua

\item\label{rulegrp_lo} regra: \texttt{1lô} -- camelô

\item\label{rulegrp_co} regra: \texttt{1cô} -- recôncavo

\item\label{rulegrp_bo} regra: \texttt{1bô}, \texttt{1dô}, \texttt{1fô}, \texttt{1gô}, \texttt{1pô}, \texttt{1mô}, \texttt{1nô}, \texttt{1rô}, \texttt{1tô}, \texttt{1vô}, \texttt{1xô}, \texttt{1zô} -- robô, judô,  telefônica, xangô, capô, sumô, econômico, tarô, chatô, vovô, saxônia, amazônia

\item\label{rulegrp_cco} regra: \texttt{1çô} -- maçônico

% As sequências gráficas ia, ie, io, oa, ua quando em posição
% final átona, que são normalmente pronunciadas como ditongos
% crescentes, mas que podem corresponder foneticamente a duas
% vogais (hiato), sobretudo quando pronunciadas pausadamente:
% vitó-ria, espé-cie, exercí-cio, nó-doa, lé-gua, tê-nue, ambí-guo.
    \setcounter{numberRulesGroups}{\value{enumi}}
\end{enumerate}

The \NumberOfNewRules{} rules were grouped above in a list of
\arabic{numberRulesGroups} types of rules. They may be further organized into four
large groups. The first, which comprises rules \ref{rulegrp_psi} to
\ref{rulegrp_zz}, includes consonant clusters such as \emph{czar}, \emph{ptose}
and \emph{gnomo}. They, unlike the examples that will be exposed in section
\ref{secxxx} (exceptions - item 9 currently), present a set of derived words,
which makes their marking advantageous in view of the number of cases that are
included in this marking rule. The second group, comprising rules
\ref{rulegrp_sub} to \ref{rulegrp_eim}, delimits the morphological boundary
between prefixes and radicals. As noted, although phonological issues guide the
separation of numerous words in Portuguese, there are also those that are
guided by morphology. This is the case of words that have the prefixes
\emph{sub-} and \emph{re-}, such as \emph{sublunar} and \emph{reinserção}. The
third group, comprising rules \ref{rulegrp_aa} to \ref{rulegrp_ia}, seeks to
understand a set of words that have vowel combinations that do not follow the
general rules. This is because the Portuguese language has vowel encounters
with the second vowel graphically marked that can be separated, forming
hiatuses, such as \emph{caótico}, \emph{balneário} and \emph{razoável}, while
there are also words with a similar structure that constitute a diphthong, such
as português, \emph{alguém} and \emph{linguística}. It is remarkable, of
course, that the latter are formed by the digraphs \emph{qu-} and \emph{gu-},
while the former by vowels other than \emph{i} and \emph{u}. The fourth and
last group, in turn, which comprises rules \ref{rulegrp_lo} to
\ref{rulegrp_cco}, which are counterparts of rules that were already in the
default rules, but did not contemplate the cases with certain accents. They
were then added to encompass words such as \emph{camelô}, \emph{recôncavo},
\emph{amazônia}, and \emph{maçônico}.

It is important to highlight that the words included in these rules are, in general, of low frequency and were incorporated into the Portuguese language without a phonotactic adaptation, which causes these idiosyncrasies and exceptions to the language. They are characterized as marked cases, since it is not possible to defend that they reproduce a phonological pattern of the language because they are not productive, that is, they are not taken as an example and/or derive new words.
